### 2008-12-17 talk/ecoop2009/benchmarks.tex
steal more text

\anto{maybe we should move this section somewhere else, if we want to use TLC
  as a running example in other sections}

In this section, we will briefly describe \emph{TLC}, a simple dynamic
language that we developed to exercise our JIT compiler generator.  As most of
dynamic languages around, \emph{TLC} is implemented through a virtual machine
that interprets a custom bytecode. Since our main interest is in the runtime
performances of the VM, we did not implement the parser nor the bytecode
compiler, but only the VM itself.

@@

Objects represent a collection of named attributes (much like Javascript or
SELF) and named methods.  At creation time, it is necessary to specify the set
of attributes of the object, as well as its methods.  Once the object has been
created, it is not possible to add/remove attributes and methods.

@@

Obviously, not all the operations are applicable to all objects. For example,
it is not possibile to \lstinline{ADD} an integer and an object, or reading an
attribute from an object which does not provide it.  Being a dynamic language,
the VM needs to do all these checks at runtime; in case one of the check
fails, the execution is simply aborted.

\anto{should we try to invent a syntax for TLC and provide some examples?}

@@

@@

\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
  on top of the evaluation stack.  As a consequence programs spend a lot of
  time pushing and popping values to/from the stack, or doing other stack
  related operations.  However, thanks ot its semplicity this is still the
  most common and preferred way to implement VMs.

\item \textbf{Boxed integers}: integer objects are internally represented as
  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
  contains the real value.  By having boxed integers, common ariithmetic
  operations are made very slow, because each time we want to load/store their
  value we need to go through an extra level of indirection.  Moreover, in
  case of a complex expression, it is necessary to create a lot of temporary
  objects to hold intermediate results.

>>>>

\anto{maybe we should move this section somewhere else, if we want to use TLC
  as a running example in other sections}
\cfbolz{yes please! how about making it the next section after the
introduction?}

In this section, we will briefly describe \emph{TLC}, a simple dynamic
language that we developed to exercise our JIT compiler generator.  As most of
dynamic languages around, \emph{TLC} is implemented through a virtual machine
that interprets a custom bytecode. Since our main interest is in the runtime
performance of the VM, we did not implement the parser nor the bytecode
compiler, but only the VM itself.

@@

Objects represent a collection of named attributes (much like JavaScript or
Self) and named methods.  At creation time, it is necessary to specify the set
of attributes of the object, as well as its methods.  Once the object has been
created, it is not possible to add/remove attributes and methods.

@@

Obviously, not all the operations are applicable to all objects. For example,
it is not possible to \lstinline{ADD} an integer and an object, or reading an
attribute from an object which does not provide it.  Being a dynamic language,
the VM needs to do all these checks at runtime; in case one of the check
fails, the execution is simply aborted.

\anto{should we try to invent a syntax for TLC and provide some examples?}
\cfbolz{we should provide an example with the assembler syntax}

@@

\cfbolz{I think this should go to the beginning of the description of the TLC as
it explains why it is written as it is written}

@@

\item \textbf{Stack based VM}: this kind of VM requires all the operands to be
  on top of the evaluation stack.  As a consequence programs spend a lot of
  time pushing and popping values to/from the stack, or doing other stack
  related operations.  However, thanks to its simplicity this is still the
  most common and preferred way to implement VMs.

\item \textbf{Boxed integers}: integer objects are internally represented as
  an instance of the \lstinline{IntObj} class, whose field \lstinline{value}
  contains the real value.  By having boxed integers, common arithmetic
  operations are made very slow, because each time we want to load/store their
  value we need to go through an extra level of indirection.  Moreover, in
  case of a complex expression, it is necessary to create many temporary
  objects to hold intermediate results.

### 2008-12-17 talk/ecoop2009/jitgen.tex
steal more text

The original idea is by Futamura \cite{Futamura99}. He proposed to generate compilers
from interpreters with automatic specialization, but his work has had
relatively little practical impact so far.

@@

Assume the Python bytecode to be constant, and constant-propagate it into the
Python interpreter.
\cfbolz{note to self: steal bits from the master thesis?}

@@

\subsection{Execution steps}

@@

\begin{itemize}
\item \emph{Green} variables contain values that are known at compile-time;
\item \emph{Red} variables contain values that are not known until run-time.
\end{itemize}

The binding-time analyzer of our translation tool-chain is based on the
same type inference engine that is used on the source RPython program,
the annotator.  In this mode, it is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.

@@

The crucial need-oriented hint is \texttt{v2 = hint(v1, concrete=True)}
which should be used in places where the programmer considers the
knowledge of the value to be essential.  This hint is interpreted by
the hint-annotator as a request for both \texttt{v1} and \texttt{v2} to be green.  It
has a \emph{global} effect on the binding times: it means that not only
\texttt{v1} but all the values that \texttt{v1} depends on – recursively –
are forced to be green.  The hint-annotator complains if the
dependencies of \texttt{v1} include a value that cannot be green, like
a value read out of a field of a non-immutable structure.

@@

In our context, though, such an error can be corrected.  This is done by
promoting a well-chosen variable among the ones that \texttt{v1} depends on.

>>>>

@@

In 1971 Yoshihiko Futamura published a paper \cite{Futamura99} that proposed a
technique to automatically transform an interpreter of a programming language
into a compiler for the same language. This would solve the problem of having to
write a compiler instead of a much simpler interpreter. He proposed to use
partial evaluation to achieve this goal. He defined partial evaluation along the following lines:

Given a program $P$ with $m + n$ input variables $s_1, ..., s_m$ and $d_1, ...,
d_n$, the partial evaluation of $P$ with respect to concrete values $s'_1, ...,
s'_m$ for the first $m$ variables is a program $P'$. The program $P'$ takes only
the input variables $d_1, ..., d_n$ but behaves exactly like $P$ with the
concrete values (but is hopefully more efficient). This transformation is done
by a program $S$, the partial evaluator, which takes $P$ and $s_1, ..., s_m$ as
input:

    $$S(P, (s'_1, ..., s'_m)) = P'$$

The variables $s_1, ..., s_m$ are called the \emph{static} variables, the
variables $d_1, ..., d_n$ are called the \emph{dynamic} variables; $P'$ is the
\emph{residual code}. Partial evaluation creates a version of $P$ that works
only for a fixed set of inputs for the first $m$ arguments. This effect is
called \emph{specialization}.

When $P$ is an interpreter for a programming language, then the $s_1, ..., s_m$
are chosen such that they represent the program that the interpreter is
interpreting and the $d_1, ..., d_n$ represent the input of this program. Then
$P'$ can be regarded as a compiled version of the program that the chosen $s'_1,
..., s'_m$ represent, since it is a version of the interpreter that can only
interpret this program. Now once the partial evaluator $S$ is implemented, it is
actually enough to implement an interpreter for a new language and use $S$
together with this interpreter to compile programs in that new language.

A valid implementation for $S$ would be to just put the concrete values into $P$
to get $P'$, which would not actually produce any performance benefits compared with
directly using $P$. A good implementation for $S$ should instead make use of the
information it has and evaluate all the parts of the program that actually
depend only on the $s_1, ..., s_m$ and to remove parts of $P$ that cannot be
reached given the concrete values.

@@

XXX write something about the problems of classical PE?

\subsection{Partial Evaluation in PyPy}

@@

\begin{itemize}
\item \emph{Green} variables contain values that are known at compile-time.
They correspond to static arguments.
\item \emph{Red} variables contain values that are usually not known
compile-time. They correspond to dynamic arguments.
\end{itemize}

The binding-time analyzer of our translation tool-chain is using a simple
abstract-interpretation based analysis. It is based on the
same type inference engine that is used on the source RPython program,
the annotator.  In this mode, it is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.
XXX the above needs rewriting when the background section is there

@@

The crucial need-oriented hint is \texttt{v2 = hint(v1, concrete=True)}
which should be used in places where the programmer considers the
knowledge of the value to be essential.  This hint is interpreted by
the hint-annotator as a request for both \texttt{v1} and \texttt{v2} to be green.  It
has a \emph{global} effect on the binding times: it means that not only
\texttt{v1} but all the values that \texttt{v1} depends on – recursively –
are forced to be green.  The hint-annotator gives an error if the
dependencies of \texttt{v1} include a value that cannot be green, like
a value read out of a field of a non-immutable instance.

@@

### 2008-12-17 talk/ecoop2009/rainbow.tex
make comments stand out more, visually
steal more text

Here are summarized the various phases of the JIT:

Translation time:

  * Low-level flowgraphs are produced

  * The *hint-annotator* colors the variables

  * The *rainbow codewriter* translates flowgraphs into rainbow bytecode

Compile-time:

  * The rainbow interpreter executes the bytecode

  * As a result, it produces executable code

Runtime:

  * The produced code is executed

@@

\subsection{Promotion}

@@

  * (compile time): the rainbow interpreter produces machine code until it
    hits a promotion point; e.g.::

@@

  * (compile time): at this point, it generates special machine code that when
    reached calls the JIT compiler again; the JIT compilation stops::

@@

  * (runtime): the machine code is executed; when it reaches a promotion
    point, it executes the special machine code we described in the previous
    point; the JIT compiler is invoked again;

  * (compile time): now we finally know the exact value of our red variable,
    and we can promote it to green; suppose that the value of 'y' is 32::

@@

  * (runtime) the execution restart from the point it stopped, until a new
    unhandled promotion point is reached.

\subsection{Virtuals and virtualizables}

>>>>

\cfbolz{XXX I think we should be very careful with the rainbow interp. it is a
total implementation-detail and we should only describe it as little as
possible}

@@

\section{Promotion}

In the sequel, we describe in more details one of the main new
techniques introduced in our approach, which we call \emph{promotion}.  In
short, it allows an arbitrary run-time value to be turned into a
compile-time value at any point in time.  Promotion is thus the central way by
which we make use of the fact that the JIT is running interleaved with actual
program execution. Each promotion point is explicitly defined with a hint that
must be put in the source code of the interpreter.

From a partial evaluation point of view, promotion is the converse of
the operation generally known as "lift" \cite{XXX}.  Lifting a value means
copying a variable whose binding time is compile-time into a variable
whose binding time is run-time – it corresponds to the compiler
"forgetting" a particular value that it knew about.  By contrast,
promotion is a way for the compiler to gain \emph{more} information about
the run-time execution of a program. Clearly, this requires
fine-grained feedback from run-time to compile-time, thus a
dynamic setting.

Promotion requires interleaving compile-time and run-time phases,
otherwise the compiler can only use information that is known ahead of
time. It is impossible in the "classical" approaches to partial
evaluation, in which the compiler always runs fully ahead of execution
This is a problem in many large use cases.  For example, in an
interpreter for a dynamic language, there is mostly no information
that can be clearly and statically used by the compiler before any
code has run.

A very different point of view on promotion is as a generalization of
techniques that already exist in dynamic compilers as found in modern
object-oriented language virtual machines.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{XXX}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results.

In the presence of promotion, dispatch optimization can usually be
reframed as a partial evaluation task.  Indeed, if the type of the
object being dispatched to is known at compile-time, the lookup can be
folded, and only a (possibly inlined) direct call remains in the
generated code.  In the case where the type of the object is not known
at compile-time, it can first be read at run-time out of the object and
promoted to compile-time.  As we will see in the sequel, this produces
very similar machine code \footnote{This can also be seen as a generalization of
a partial evaluation transformation called "The Trick" (see e.g. \cite{XXX}),
which again produces similar code but which is only applicable for finite sets
of values.}.

The essential advantage is that it is no longer tied to the details of
the dispatch semantics of the language being interpreted, but applies in
more general situations.  Promotion is thus the central enabling
primitive to make partial evaluation a practical approach to language
independent dynamic compiler generation.

\subsection{Promotion as Applied to the TLC}

XXX

\subsection{Promotion in Practise}

@@

\begin{itemize}
  \item (compile time): the rainbow interpreter produces machine code until it
    hits a promotion point; e.g.::

@@

  \item (compile time): at this point, it generates special machine code that when
    reached calls the JIT compiler again; the JIT compilation stops::

@@

  \item (runtime): the machine code is executed; when it reaches a promotion
    point, it executes the special machine code we described in the previous
    point; the JIT compiler is invoked again;

  \item (compile time): now we finally know the exact value of our red variable,
    and we can promote it to green; suppose that the value of 'y' is 32::

@@

  \item (runtime) the execution restart from the point it stopped, until a new
    unhandled promotion point is reached.
\end{itemize}

\section{Automatic Unboxing of Intermediate Results}

XXX the following section needs a rewriting to be much more high-level and to
compare more directly with classical escape analysis

Interpreters for dynamic languages typically allocate a lot of small
objects, for example due to boxing.  For this reason, we
implemented a way for the compiler to generate residual memory
allocations as lazily as possible.  The idea is to try to keep new
run-time structures "exploded": instead of a single run-time pointer to
a heap-allocated data structure, the structure is "virtualized" as a set
of fresh variables, one per field.  In the compiler, the variable that
would normally contain the pointer to the structure gets instead a
content that is neither a run-time value nor a compile-time constant,
but a special \emph{virtual structure} – a compile-time data structure that
recursively contains new variables, each of which can again store a
run-time, a compile-time, or a virtual structure value.

This approach is based on the fact that the "run-time values" carried
around by the compiler really represent run-time locations – the name of
a CPU register or a position in the machine stack frame.  This is the
case for both regular variables and the fields of virtual structures.
It means that the compilation of a \texttt{getfield} or \texttt{setfield}
operation performed on a virtual structure simply loads or stores such a
location reference into the virtual structure; the actual value is not
copied around at run-time.

It is not always possible to keep structures virtual.  The main
situation in which it needs to be "forced" (i.e. actually allocated at
run-time) is when the pointer escapes to some non-virtual location like
a field of a real heap structure.

Virtual structures still avoid the run-time allocation of most
short-lived objects, even in non-trivial situations.  The following
example shows a typical case.  Consider the Python expression \texttt{a+b+c}.
Assume that \texttt{a} contains an integer.  The PyPy Python interpreter
implements application-level integers as boxes – instances of a
\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
addition of two integers:

XXX needs to use TLC examples
\begin{verbatim}
    def add(w1, w2):            # w1, w2 are W_IntObject instances
        value1 = w1.intval
        value2 = w2.intval
        result = value1 + value2
        return W_IntObject(result)
\end{verbatim}

When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
issued; the intermediate \texttt{W\_IntObject} instance is built by the first
call and thrown away after the second call.  By contrast, when the
interpreter is turned into a compiler, the construction of the
\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
field directly references the register in which the run-time addition
put its result.  This location is read out of the virtual structure at
the beginning of the second \texttt{add()}, and the second run-time addition
directly operates on the same register.

An interesting effect of virtual structures is that they play nicely with
promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
function for integers, it must first determine that the two arguments
are indeed integer objects.  In the corresponding dispatch logic, we
have added two hints to promote the type of each of the two arguments.
This produces a compiler that has the following behavior: in the general
case, the expression \texttt{a+b} will generate two consecutive run-time
switches followed by the residual code of the proper version of
\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
the intermediate value will contain a compile-time constant as type.
Promoting a compile-time constant is trivial – no run-time code is
generated.  The whole expression \texttt{a+b+c} thus only requires three
switches instead of four.  It is easy to see that even more switches can
be skipped in larger examples; typically, in a tight loop manipulating
only integers, all objects are virtual structures for the compiler and
the residual code is theoretically optimal – all type propagation and
boxing/unboxing occurs at compile-time.
